### 2022-08-08 main/chapter_2ljets.tex
begin writing on odd discovery region results

\paragraph{Discovery}
\TODO{pruning $1\%$}

Figure~\ref{fig:2ljets_test_stat_distributions}

Figure~\ref{fig:2ljets_test_stat_distributions}

\begin{figure}[tp]
\centering
\begin{subfigure}{0.48\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/2ljets_upper_limit_scan_drhigh.pdf}
\caption{DR-High}
\end{subfigure}
\hfill
\begin{subfigure}{0.48\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/2ljets_upper_limit_scan_drlow.pdf}
\caption{DR-Low}
\end{subfigure}
\caption[
Upper limit scans for DR-High and DR-Low
]{%
Upper limit scans for (a) DR-High and (a) DR-Low.
Solid black lines interpolate the $p$-values $\clsb$ and $\clb$ estimated at
each marked point with the observed data.
The upper limit is reported where the interpolation of their ratio $\cls$
crosses $0.05$.
Coloured backgrounds illustrate what would have been observed with data at
$\pm1$-sigma quantiles of the background-only distribution.
Each point uses $500$k ``toy'' samples for the signal-plus-background
distribution, and $1$m samples for the backgroun-only distribution.
Individual distributions are inspected in
Figure~\ref{fig:2ljets_test_stat_distributions}.
}
\label{fig:2ljets_upper_limit_scans}
\end{figure}

\begin{figure}[tp]
\centering
\begin{subfigure}{0.50\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/2ljets_test_stat_distribution_drhigh3.pdf}
\caption{DR-High, $\mu_\textrm{Discovery}=3$}
\end{subfigure}
\hfill
\begin{subfigure}{0.46\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/2ljets_test_stat_distribution_drlow10.pdf}
\caption{DR-Low, $\mu_\textrm{Discovery}=10$}
\end{subfigure}
\caption[
Test statistic distributions for points in the DR-High and DR-Low scans
]{%
Test statistic distributions for points in the (a) DR-High and (b) DR-Low
scans, with $\mu_\textrm{Discovery}=3$ and $\mu_\textrm{Discovery}=10$
respectively.
Axis limits are set from the minimum and maximum test statistics recorded.
Note the different binning: histograms contain normalized densisites with
different widths between the two distributions.
The black vertical line indicates the observed test statistic,
and the $p$-values $\clb$ and $\clsb$ are tail areas to its right.
If the two distributions were predictive, their relative densities at that
observation would indicate their likelihood ratio.
Results are:
% clb =  0.001322
% clsb =  5.7e-05
% cls =  0.0431164901664
(a, DR-High) $\clb = 1.3\times 10^{-3}$ and $\clsb = 5.7\times 10^{-5}$,
% clb =  0.087502
% clsb =  0.007667
% cls =  0.0876208543805
(b, DR-Low) $\clb = 8.8\times 10^{-2}$ and $\clsb = 7.7\times 10^{-3}$.
}
\label{fig:2ljets_test_stat_distributions}
\end{figure}

>>>>

\paragraph{Discovery}
Our discovery region results,
which have been presented in Table~\ref{tab:2ljets_discovery},
feature, and veil, some interesting quirks.

\begin{figure}[tp]
\centering
\begin{subfigure}{0.48\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/2ljets_upper_limit_scan_drhigh.pdf}
\caption{DR-High}
\end{subfigure}
\hfill
\begin{subfigure}{0.48\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/2ljets_upper_limit_scan_drlow.pdf}
\caption{DR-Low}
\end{subfigure}
\\[.5ex]
\begin{subfigure}{0.48\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/2ljets_upper_limit_scan_droffshell.pdf}
\caption{DR-OffShell}
\end{subfigure}
\caption[
Upper limit scans for DR-High and DR-Low
]{%
Upper limit scans for (a) DR-High, (b) DR-Low, and (c) DR-OffShell.
Solid black lines interpolate the $p$-values $\clsb$ and $\clb$ estimated at
each marked point with the observed data.
The upper limit is reported where the interpolation of their ratio $\cls$
crosses $0.05$.
\\
Coloured backgrounds illustrate what would have been observed with data at
$\pm1$-sigma quantiles of the background-only distribution.
Each point uses $500$k ``toy'' samples for the background-only
distribution, and $1$m samples for the signal-plus-background distribution.
Individual distributions are inspected in
Figure~\ref{fig:2ljets_test_stat_distributions}.
\\
\emph{These figures were produced with \histfitter\ tools through
my ``Bailed statistics'' software~\cite{bailedstatisticsgithub}.}
}
\label{fig:2ljets_upper_limit_scans}
\end{figure}

Perhaps what should be the most meaningful result is the observed upper limit
$S_{\mathrm{obs}}^{95}$.
The analysis of each region sets its upper limit by scanning different values
of the signal expectation $\mu_\textrm{Discovery}$ and interpolating to
estimate the point where $\cls(\mu_\textrm{Discovery}^*) = 0.05$.
That interpolated point determines
$\mu_\textrm{Discovery}^* = S_{\mathrm{obs}}^{95}$, and the reported value
of $\clb$ is also reported at that signal expectation
$\mu_\textrm{Discovery}^*$, since the test statistic is a function of
$\mu_\textrm{Discovery}$.
Three examples of these scans are visualized in
Figure~\ref{fig:2ljets_upper_limit_scans}.

Some of our discovery regions have small yields for which the asymptotic
approximations~\cite{Cowan:2010js} to test statistic distributions are not
believed to work.
For consistency, we therefore use the alternative toy simulation
approximation for all discovery regions.
Sampling of the test statistic distributions
(with \histfitter-\roofit\ software)
is extremely computationally expensive, so cut those costs by pruning our
discovery regions models \histfitter's ``method 2'' at a $1\%$ tolerance.
As discussed previously, this makes negligible differences to the results.
The exact pruned models are shared in our \hepdata\
archive~\cite{hepdata.116034}.
Not only do these simulations take computational time, they also suffer
torrential memory leaks that halt progress beyond small sample sizes.
To work around this, we evaluate and combine many small samples, with different
random seeds, with software that I name
``Bailed statistics''~\cite{bailedstatisticsgithub}%
\footnote{%
To share software that assists in spending computational resources to
calculate statistics that are so deeply flawed causes me some shame.
In a vain attempt to absolve myself of this sin, and to positively encourage
honest reporting of the data, I include with it the following note:
\emph{%
``Hypotheses compare through the relative likelihoods they assign to data.
A p-value is a cumulative distribution function evaluated at data; CLs is a
ratio of p-values.
Please present results of this software accurately and clearly.
Examples of false or unclear presentations of a p-value or CLs include:
--- as a probability that an hypothesis is true or false,
--- as a probability of compatibility with an hypothesis,
--- as a probability that data occurred at random, by chance, or by statistical
fluctuation,
--- as a likelihood or importance of data,
--- as necessary for an optimal or rational decision rule.
Please also respect that the association of a p-value or CLs with the words
`test', `limit', `confidence', `significance', `exclusion', `evidence',
`observation' or `discovery' is nominal,
and may not reflect the words' meanings in English.''
}%
}
as it bails the leaks overboard.
Despite this workaround, the simulations took several weeks of parallel
processing on around $1500$ processor cores, accumulating to around a million
core hours.

\begin{figure}[tp]
\centering
\begin{subfigure}{0.50\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/2ljets_test_stat_distribution_drhigh3.pdf}
\caption{DR-High, $\mu_\textrm{Discovery}=3$}
\label{fig:2ljets_test_stat_distributions_drhigh}
\end{subfigure}
\hfill
\begin{subfigure}{0.46\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/2ljets_test_stat_distribution_drlow10.pdf}
\caption{DR-Low, $\mu_\textrm{Discovery}=10$}
\end{subfigure}
\\[.5ex]
\begin{subfigure}{0.46\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/2ljets_test_stat_distribution_droffshell14.pdf}
\caption{DR-OffShell, $\mu_\textrm{Discovery}=14$}
\end{subfigure}
\caption[
Test statistic distributions for points in the DR-High and DR-Low scans
]{%
Test statistic distributions for points in the (a) DR-High, (b) DR-Low,
and (c) DR-OffShell scans at points close to their reported upper limits:
$\mu_\textrm{Discovery}=3$, $\mu_\textrm{Discovery}=10$,
and $\mu_\textrm{Discovery}=14$ respectively.
\\
Axis limits are set from the minimum and maximum test statistics recorded.
Note the different binning: histograms contain normalized densities with
different widths between the two distributions.
The black vertical line indicates the observed test statistic,
and the $p$-values $\clb$ and $\clsb$ are tail areas to its right.
If the two distributions were predictive, their relative densities at that
observation would indicate their likelihood ratio.
\\
Results are:
% clb =  0.001322
% clsb =  5.7e-05
% cls =  0.0431164901664
(a) $\clb = 1.3\times 10^{-3}$ and $\clsb = 5.7\times 10^{-5}$,
% clb =  0.087502
% clsb =  0.007667
% cls =  0.0876208543805
(b) $\clb = 8.8\times 10^{-2}$ and $\clsb = 7.7\times 10^{-3}$,
% clb =  0.67508
% clsb =  0.03661
% cls =  0.0542306097055
and (c) $\clb = 0.68$ and $\clsb = 0.037$.
\\
\emph{These figures were adapted from plots produced with \histfitter\ tools
through my ``Bailed statistics'' software~\cite{bailedstatisticsgithub}.}
}
\label{fig:2ljets_test_stat_distributions}
\end{figure}

Many test statistic distributions are involved here --- two (background-only
and signal-plus-background) at each scanned point, and one additional
distribution per region (with yet another test statistic) for the discovery
$p$-value $p_{s\!=\!0}$.
From these, a few examples are presented in
Figure~\ref{fig:2ljets_test_stat_distributions}
to investigate curious features of the results.

Our first curiosity is that the reported $S_{\mathrm{obs}}^{95} = 3.0$
for DR-High is rounded up from a result of $2.93$ that was evaluated with these
simulation methods.
This makes sense --- the limit for $n=0$ should arguably always be $3$ since
if a signal adds $\mu$ events, then the Poisson likelihood for zero data always
scales by $e^{-\mu} \approx 0.05$, and so any likelihood ratio between
background-only and signal-plus-background models is $0.05$ for $\mu = 3$.
By taking a ratio of $p$-values, $\cls$ approximates that likelihood ratio:
we see from Figure~\ref{fig:2ljets_test_stat_distributions_drhigh} that the
observed data does land in a spike at the end of the distribution that
corresponds to to pseudo-data with $n=0$; the peaks to its left correspond
to $n=1$, $n=2$ and so on, with widths due to the other parameters varied in
toy sampling.

The $n=0$ case has no visible width since its likelihood is linear
(adding signal just adds a constant to the $-\log$ likelihood objectives being
minimized), but its distribution is not quite a delta function since the test
statistic depends on the noisy termination of numerical optimization
algorithms.
Resulting $\clb$ and $\clsb$ values depend on this effect ---
with $3.9$ expected background events and zero data we expect
$\clb = \mathrm{Poisson}(0\mid 3.9) = e^{-3.9} \approx 2.0\times 10^{-2}$.
And that is $15$ times larger than the evaluated $\clb = 1.3\times 10^{-3}$
because, as shown in Figure~\ref{fig:2ljets_test_stat_distributions_drhigh_n0},
the observed test statistic happened to fall quite far to the right
of that $n=0$ noise distribution.
This would appear to be an anomalously large significance if $\clb$ were
misinterpreted as a measure of compatibility with the background model.
Thankfully it is remedied by the $\cls$-prescribed ratio with $\clsb$,
since both $p$-values suffer from similar effects.
Statistical precision is harmed, however --- although $1$m samples should
give us around $e^{-7.9} \times 10^6 \approx 1000$ signal samples with
$n=0$, only $57$ toy samples land to the right of the observed test statistic.
And with $\sqrt{n}$ precision, that easily explains the difference between
limits at $3.0$ and $2.93$ despite the huge sample size.

\TODO{}
% real world costs to pretending that our uncertainties are auxiliary data
% not only idle Bayesian-versus-Frequentist philosophy
% other examples - cite 4.6 limit.

\begin{figure}[tp]
\centering
\includegraphics[width=0.6\textwidth]{figures/2ljets_drhigh_plot_n0.pdf}
\caption[
Test statistic distribution for DR-High close to 3
]{%
Test statistic distribution for DR-High close to $3$.
This zooms in to the narrow spike of toy samples that should have $n=0$
observed data in the discovery region.
Because the likelihood function for zero data,
$\mathrm{Poisson}(0\mid x) = e^{-x}$, is $\log$-linear,
this would analytically be a delta spike, but it acquires width from the finite
precision of numerical optimization algorithms.
The observed test statistic falling far to the right of that numerical noise
reduces our statistical precision in estimating $\clb$ and $\clsb$.
}
\label{fig:2ljets_test_stat_distributions_drhigh_n0}
\end{figure}
